Open archaeology, open source?

### 2024-02-06 analysis/_04-open_archaeology.qmd
Adjust range of fig-github-cumulative

p_gh_commits_cum <- oarch |>
  unnest(gh_commits) |>
  drop_na(sha) |>
  mutate(date = floor_date(datetime, "week")) |>
  group_by(category, date) |>
  summarise(n_commits = n(), .groups = "drop_last") |>
  mutate(cum_commits = cumsum(n_commits)) |>
  group_by(category) |>
  nest() |>
  mutate(data = map(data, \(x) { # Extend the range of each category to end
    bind_rows(x, list(date = dt_last_commit, cum_commits = max(x$cum_commits)))
  })) |>
  unnest(c(data)) |>
  ggplot(aes(date, cum_commits,
             fill = fct_rev(category),
             colour = fct_rev(category))) +
  geom_area() +
  annotate("segment", 
           x = ymd("2008-04-01", tz = "UTC"), 
           xend = ymd("2008-04-01", tz = "UTC"),
           y = 16000, yend = 4000,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text", 
           x = ymd("2008-04-01", tz = "UTC"), y = 19000,
           label = "GitHub\nlaunched",
           vjust = 0) +
  scale_x_datetime(limits = c(ymd("2006-12-31", tz = "UTC"), dt_last_commit),
                   date_breaks = "2 years", 
                   date_labels = "%Y",
                   expand = c(0,0)) +
  scale_y_continuous(labels = scales::comma) +
  scale_colour_bright(guide = guide_none()) +
  scale_fill_bright() +
  labs(
    y = "Cumulative commits",
    x = NULL, fill = NULL
  ) +
  theme_oarch() +
  theme(legend.position = "top")

p_gh_repos_cum <- oarch |>
  unnest(gh_commits) |>
  drop_na(sha) |>
  group_by(repo) |> 
  filter(datetime == min(datetime)) |> # Only first commits
  mutate(date = floor_date(datetime, "week")) |>
  group_by(category, date) |>
  summarise(n_repos = n(), .groups = "drop_last") |>
  mutate(cat_total = sum(n_repos)) |>
  mutate(cum_repos = cumsum(n_repos)) |>
  select(category, date, cum_repos) |>
  group_by(category) |>
  nest() |>
  mutate(data = map(data, \(x) { # Extend the range of each category to end
    bind_rows(x, list(date = dt_last_commit, cum_repos = max(x$cum_repos)))
  })) |>
  unnest(c(data)) |>
  ggplot(aes(date, cum_repos,
             fill = fct_rev(category),
             colour = fct_rev(category))) +
  geom_area() +
  annotate("segment",
           x = ymd("2008-04-01", tz = "UTC"),
           xend = ymd("2008-04-01", tz = "UTC"),
           y = 40, yend = 10,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text",
           x = ymd("2008-04-01", tz = "UTC"), y = 50,
           label = "GitHub\nlaunched",
           vjust = 0) +
  scale_x_datetime(limits = c(ymd("2006-12-31", tz = "UTC"), dt_last_commit),
                   date_breaks = "2 years", 
                   date_labels = "%Y",
                   expand = c(0,0)) +
  scale_y_continuous(labels = scales::comma) +
  scale_colour_bright(guide = guide_none()) +
  scale_fill_bright(guide = guide_none()) +
  labs(
    y = "Cumulative repositories",
    x = NULL, fill = NULL
  ) +
  theme_oarch()

@@

As we can see <!-- this is no longer evident in the figure -->, archaeologists have been using git from even before GitHub was launched in 2008.
But use of git really began to take off around 2014--2015, when we see an uptick in the rate of growth.
Around this time we also see that GitHub starts being used to host documents and scripts.
This may represent a recognition of GitHub's ability to track things other than code, and a willingness to experiment with version control systems as a medium for disseminating work in an open and somewhat nerdy way.

>>>>

p_gh_commits_cum <- oarch |>
  unnest(gh_commits) |>
  drop_na(sha) |>
  mutate(date = floor_date(datetime, "week")) |>
  group_by(category, date) |>
  summarise(n_commits = n(), .groups = "drop_last") |>
  mutate(cum_commits = cumsum(n_commits)) |>
  group_by(category) |>
  nest() |>
  mutate(data = map(data, \(x) { # Extend the range of each category to end
    bind_rows(x, list(date = dt_last_commit, cum_commits = max(x$cum_commits)))
  })) |>
  unnest(c(data)) |>
  ggplot(aes(date, cum_commits,
             fill = fct_rev(category),
             colour = fct_rev(category))) +
  geom_area() +
  annotate("segment", 
           x = ymd("2008-04-01", tz = "UTC"), 
           xend = ymd("2008-04-01", tz = "UTC"),
           y = 16000, yend = 4000,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text", 
           x = ymd("2008-04-01", tz = "UTC"), y = 19000,
           label = "GitHub\nlaunched",
           vjust = 0) +
  scale_x_datetime(limirs = c(ymd("2004-12-31", tz = "UTC"), dt_last_commit),
                   date_breaks = "2 years", 
                   date_labels = "%Y",
                   expand = c(0,0)) +
  scale_y_continuous(labels = scales::comma) +
  scale_colour_bright(guide = guide_none()) +
  scale_fill_bright() +
  labs(
    y = "Cumulative commits",
    x = NULL, fill = NULL
  ) +
  theme_oarch() +
  theme(legend.position = "top")

p_gh_repos_cum <- oarch |>
  unnest(gh_commits) |>
  drop_na(sha) |>
  group_by(repo) |> 
  filter(datetime == min(datetime)) |> # Only first commits
  mutate(date = floor_date(datetime, "week")) |>
  group_by(category, date) |>
  summarise(n_repos = n(), .groups = "drop_last") |>
  mutate(cat_total = sum(n_repos)) |>
  mutate(cum_repos = cumsum(n_repos)) |>
  select(category, date, cum_repos) |>
  group_by(category) |>
  nest() |>
  mutate(data = map(data, \(x) { # Extend the range of each category to end
    bind_rows(x, list(date = dt_last_commit, cum_repos = max(x$cum_repos)))
  })) |>
  unnest(c(data)) |>
  ggplot(aes(date, cum_repos,
             fill = fct_rev(category),
             colour = fct_rev(category))) +
  geom_area() +
  annotate("segment",
           x = ymd("2008-04-01", tz = "UTC"),
           xend = ymd("2008-04-01", tz = "UTC"),
           y = 40, yend = 10,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text",
           x = ymd("2008-04-01", tz = "UTC"), y = 50,
           label = "GitHub\nlaunched",
           vjust = 0) +
  scale_x_datetime(limits = c(ymd("2004-12-31", tz = "UTC"), dt_last_commit),
                   date_breaks = "2 years", 
                   date_labels = "%Y",
                   expand = c(0,0)) +
  scale_y_continuous(labels = scales::comma) +
  scale_colour_bright(guide = guide_none()) +
  scale_fill_bright(guide = guide_none()) +
  labs(
    y = "Cumulative repositories",
    x = NULL, fill = NULL
  ) +
  theme_oarch()

@@

As we can see, archaeologists have been using git from even before GitHub was launched in 2008.
But use of git really began to take off around 2014--2015, when we see an uptick in the rate of growth.
Around this time we also see that GitHub starts being used to host documents and scripts.
This may represent a recognition of GitHub's ability to track things other than code, and a willingness to experiment with version control systems as a medium for disseminating work in an open and somewhat nerdy way.
